---
title: "Histogram-based Method"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(MASS)
library(dplyr)
library(gridExtra)
library(mvtnorm)
```

## Theoretical Background

### Estimator

- Goal: estimating $w = f^{(0)} / f^{(1)}$ from $\{x_{i^{(k)}}\}^{n_k}_{i=1} \overset{\mathrm{iid}}{\sim} f^{(k)}$

- Estimator[(Kpotufe, 2017)](https://proceedings.mlr.press/v54/kpotufe17a/kpotufe17a.pdf): $\tilde{w} = \frac{\hat{\mathbb{P}}^{(0)}(B(x, r))}{\hat{\mathbb{P}}^{(1)}(B(x, r))} = \frac{n_{0}^{-1}\sum_{i=1}^{n_0}\mathbb{1}(||x - x_{i}^{(0)}||_2 \leq r)}{n_{1}^{-1}\sum_{i=1}^{n_1}\mathbb{1}(||x - x_{i}^{(1)}||_2 \leq r}$; $\hat{w}(x) = \tilde{w}(x) \mathbb{1}(\hat{\mathbb{P}}^{(1)}(B(x, r)) \geq \alpha)$

- Theorem 3.4.1 [(Kpotufe, 2017)](https://proceedings.mlr.press/v54/kpotufe17a/kpotufe17a.pdf):

  With $r$ $\asymp [\frac{log(n_0 \bigwedge n_{1})}{n_0 \bigwedge n_1}]^{\frac{\beta}{2\beta + d}}$, $\alpha \asymp \frac{log(n_0)}{n_0}$, we have:

  $\mathbb{E}_{X \sim \mathbb{P}^{(1)}} |\hat{w}(X) - w(X)| \lesssim_{\mathbb{P}}[\frac{log(n_0 \bigwedge n_{1})}{n_0 \bigwedge n_1}]^{\frac{\beta}{2\beta + d}}$

### Assumptions

- Compact Support: $supp(\mathbb{P}^{(1)})$ is compact in $\mathbb{R}^{d}$

- Bounded Density Ratio: $||f||_{\infty} \leq B < \infty$

- $\beta$-Hölder class: $|w(x) - w(x')| \leq L||x - x'||_{2}^{\beta}$ with $\beta \in (0, 1]$

## Implementation



```{r}
# ----------------------------
# 1) Setup
# ----------------------------
set.seed(123)  # for reproducibility

D <- 20
v <- seq_len(D)  # v = (1, 2, ..., 20)

n_source <- 1000
n_target <- 1000

# ----------------------------
# 2) Define mixture distributions
# ----------------------------
# Q (target): t ~ 0.5*N(-4, 1) + 0.5*N(4, 1)
# P (source): t ~ 0.5*N(-4, 4) + 0.5*N(4, 4)

# A helper function to sample from a 2-component mixture
# with equal weights 0.5 and 0.5, given:
#   means = c(m1, m2), sds = c(sd1, sd2), and total n
mix_2_gaussian <- function(n, means, sds) {
  # Step 1: Decide which component each point comes from
  comps <- sample.int(2, size = n, replace = TRUE, prob = c(0.5, 0.5))
  
  # Step 2: For each chosen component, draw from corresponding Normal
  rnorm(n, mean = means[comps], sd = sds[comps])
}

# ----------------------------
# 3) Generate the scalars t
# ----------------------------
target_t <- mix_2_gaussian(
  n    = n_target,
  means = c(-4, 4),
  sds   = c(1, 1)
)

source_t <- mix_2_gaussian(
  n    = n_source,
  means = c(-4, 4),
  sds   = c(2, 2)  # because variance=4 => sd=2
)

# ----------------------------
# 4) Embed each scalar in D-dimensional space: X = t * v
# ----------------------------
X_target <- outer(target_t, v)  # n_target x D matrix
X_source <- outer(source_t, v)  # n_source x D matrix

# Optionally, convert to data frames:
df_target <- as.data.frame(X_target)
df_source <- as.data.frame(X_source)

# Now you have (X_source, X_target) with dimension 20,
# each lying on the line spanned by v (since x = t*v).

f_target <- function(t) {
  0.5 * dnorm(t, mean = -4, sd = 1) +
  0.5 * dnorm(t, mean = 4,  sd = 1)
}

f_source <- function(t) {
  0.5 * dnorm(t, mean = -4, sd = 2) +
  0.5 * dnorm(t, mean = 4,  sd = 2)
}

# True weight function w(t)
w_fun <- function(t) {
  f_target(t) / f_source(t)
}


```

We generate multi-Gaussian distribution to simulate non-parametric data. Dimension of data $d$ is `r D`; sample size of target data $n_0$ is `r n_target`; and sample size of source data $n_1$ is `r n_source`.

All the parameters we use are from [Kpotufe(2017)](https://proceedings.mlr.press/v54/kpotufe17a/kpotufe17a.pdf).

```{r simulate_estimation}
compute_histogram_ratio <- function(x, target_data, source_data, r) {
  # x is a numeric vector of length d
  # target_data: n0 x d
  # source_data: n1 x d
  # r: numeric radius
  
  n0 <- nrow(target_data)
  n1 <- nrow(source_data)
  
  # Indicator sum for target_data
  # 1/ if distance <= r, 0/ otherwise
  # We'll compute Euclidean distance from x to each row in target_data
  dist_target <- apply(target_data, 1, function(xi) sqrt(sum((xi - x)^2)))
  indicators_target <- (dist_target <= r)
  numerator <- mean(indicators_target)  # (1/n0) sum_i 1(...)
  
  # Indicator sum for source_data
  dist_source <- apply(source_data, 1, function(xi) sqrt(sum((xi - x)^2)))
  indicators_source <- (dist_source <= r)
  denominator <- mean(indicators_source)  # (1/n1) sum_j 1(...)
  
  # Avoid division by zero if the denominator = 0
  if (denominator == 0) {
    ratio <- Inf  # or define a small offset to avoid Inf, e.g. ratio <- numerator / (denominator + 1e-10)
  } else {
    ratio <- numerator / denominator
  }
  
  # To obtain the re-weighted test vector
  alpha = log(n0) / n0
  estimate_reweight <- ifelse(
    denominator >= alpha, ratio, 0
  )
  
  list(ratio = ratio, 
       estimation = estimate_reweight)
}

target_sample_n = 1000
source_sample_n = 1000
min_sample = min(target_sample_n, source_sample_n)
# Choose parameters
r = 45
test_n = 1000
set.seed(335)

test_t <- mix_2_gaussian(
  n     = test_n,
  means = c(-4, 4),
  sds   = c(2, 2)
)

# Embed in D dimensions
X_test <- outer(test_t, v)
test_estimations <- matrix(NA, nrow = test_n, ncol = 2)
weighted_ratio   <- numeric(test_n)

for (i in seq_len(test_n)) {
  # x_test is 1 x D
  x_test <- X_test[i, ]   # a single test sample
  
  # Compute histogram ratio
  hist_ratio <- compute_histogram_ratio(x_test, X_target, X_source, r)
  weighted_ratio[i]         <- hist_ratio$ratio
  test_estimations[i, 1]    <- hist_ratio$estimation
  test_estimations[i, 2]    <- x_test[1] / v[1]  # store the 1st coordinate, or any coordinate
}

estimations_df <- data.frame(
  estimation = test_estimations[, 1],
  t          = test_estimations[, 2]
)
```

```{r}
# Visualize estimation result
df_t_w_source <- data.frame(
  t = source_t,
  w = w_fun(source_t)
)

source_target <- ggplot() +
  geom_density(data = df_target, aes(x = V1, color = "Target")) +
  geom_density(data = df_source, aes(x = V1, color = "Source")) +
  scale_color_manual(
    name   = "distribution",  # this is the legend title
    values = c("Target" = "blue",
               "Source" = "red")
  ) +
  labs(x = "t", y = "Density", title = "Target vs Source Distribution") +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 6),  # Smaller legend title
    legend.text  = element_text(size = 6)    # Smaller legend text
  )

estimation_true <- ggplot() +
  geom_point(
    data = df_t_w_source,
    aes(x = t, y = w, color = "True Weight"),
    alpha = 0.3
  ) +
  # Estimated weight (red)
  geom_point(
    data = estimations_df,
    aes(x = t, y = estimation, color = "Estimated Weight"),
    alpha = 0.3
  ) +
  scale_color_manual(
    name   = "weights",  # this is the legend title
    values = c("True Weight" = "blue",
               "Estimated Weight" = "red")
  ) +
  labs(
    x = "t",
    y = "Weight",
    title = "Estimated vs. True Density"
  ) +
  theme_minimal() +
  theme(
    legend.title = element_text(size = 7),  # Smaller legend title
    legend.text  = element_text(size = 7)    # Smaller legend text
  )

grid.arrange(source_target, estimation_true, ncol = 2)
```

## Highlights

- Histogram-based method allow us to estimate the target data's distribution with source data whether the data is parametric or not. And it doesn't suffer from the curse of high dimension.

- To find the optimal oracle rate $r$, [Kpotufe](https://proceedings.mlr.press/v54/kpotufe17a.html) suggests a data-driven procedure in the section 4.2 of his paper.

## References

[Kpotufe, S. (2017). Lipschitz density-ratios, structured data, and data-driven tuning. In Artificial Intelligence and Statistics, pages 1320–1328. PMLR.](https://proceedings.mlr.press/v54/kpotufe17a.html)
