---
title: "Histogram-based Method"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(mvtnorm)
```

## Theoretical Background

### Estimator

### Assumptions

- Compact Support: $supp(\mathbb{P}^{(1)})$ is compact in $\mathbb{R}^{d}$

- Bounded Density Ratio: $||f||_{\infty} \leq B < \infty$

- $\beta$-Hölder class: $|w(x) - w(x')| \leq L||x - x'||_{2}^{\beta}$ with $\beta \in (0, 1]$

## Implementation

```{r simulate_data}
simulate_data <- function(n1 = 1000,   # number of target samples
                          n0 = 1000,   # number of source samples
                          d  = 2,      # dimension
                          seed = 123) {
  set.seed(seed)
  
  # --- Generate Source Data (Mixture of Gaussians) ---
  p_source <- 0.5  # mixture proportion (50% from component 1, 50% from component 2)
  
  # Means for the two components in the source mixture
  mu_source1 <- rep(0, d)
  mu_source2 <- rep(3, d)  # shift second component to separate it somewhat
  
  # Covariance matrices for the two source components (identity for simplicity)
  Sigma_source1 <- diag(d)
  Sigma_source2 <- diag(d)
  
  # Allocate memory
  source_data <- matrix(NA, nrow = n1, ncol = d)
  
  # Decide which component each observation belongs to
  z_source <- rbinom(n1, size = 1, prob = p_source)
  
  for (i in seq_len(n1)) {
    if (z_source[i] == 1) {
      source_data[i, ] <- mvrnorm(n = 1, mu = mu_source1, Sigma = Sigma_source1)
    } else {
      source_data[i, ] <- mvrnorm(n = 1, mu = mu_source2, Sigma = Sigma_source2)
    }
  }
  
  # --- Generate Target Data (Another Mixture of Gaussians) ---
  p_target <- 0.7  # mixture proportion for target data
  mu_target1 <- rep(1,  d)   # slightly different mean
  mu_target2 <- rep(-2, d)   # negative shift
  Sigma_target1 <- diag(d) * 0.5  # smaller covariance
  Sigma_target2 <- diag(d) * 2    # larger covariance
  
  # Allocate memory
  target_data <- matrix(NA, nrow = n0, ncol = d)
  
  # Mixture indicator for target distribution
  z_target <- rbinom(n0, size = 1, prob = p_target)
  
  for (i in seq_len(n0)) {
    if (z_target[i] == 1) {
      target_data[i, ] <- mvrnorm(n = 1, mu = mu_target1, Sigma = Sigma_target1)
    } else {
      target_data[i, ] <- mvrnorm(n = 1, mu = mu_target2, Sigma = Sigma_target2)
    }
  }
  
  # Return both datasets
  list(
    source = source_data,
    target = target_data
  )
}

# Example Usage
target_sample_n = 800
source_sample_n = 1000
dimension = 2
sim_data <- simulate_data(n1 = source_sample_n, n0 = target_sample_n, d = dimension, seed = 123)
source_data <- sim_data$source
target_data <- sim_data$target

```

We generate multi-Gaussian distribution to simulate non-parametric data. Dimension of data $d$ is `r dimension`; sample size of target data $n_0$ is `r target_sample_n`; and sample size of source data $n_1$ is `r source_sample_n`.

```{r raw_data_visualization}
# Visualize high dimensional data
```

```{r simulate_estimation}
target_sample_n = 800
source_sample_n = 1000
min_sample = min(target_sample_n, source_sample_n)
# Choose parameters
r = 0.5731 # we may use monte-carlo simulation
alpha = log(target_sample_n) / target_sample_n


compute_histogram_ratio <- function(x, target_data, source_data, r) {
  # x is a numeric vector of length d
  # target_data: n0 x d
  # source_data: n1 x d
  # r: numeric radius
  
  n0 <- nrow(target_data)
  n1 <- nrow(source_data)
  
  # Indicator sum for target_data
  # 1/ if distance <= r, 0/ otherwise
  # We'll compute Euclidean distance from x to each row in target_data
  dist_target <- apply(target_data, 1, function(xi) sqrt(sum((xi - x)^2)))
  indicators_target <- (dist_target <= r)
  numerator <- mean(indicators_target)  # (1/n0) sum_i 1(...)
  
  # Indicator sum for source_data
  dist_source <- apply(source_data, 1, function(xi) sqrt(sum((xi - x)^2)))
  indicators_source <- (dist_source <= r)
  denominator <- mean(indicators_source)  # (1/n1) sum_j 1(...)
  
  # Avoid division by zero if the denominator = 0
  if (denominator == 0) {
    ratio <- Inf  # or define a small offset to avoid Inf, e.g. ratio <- numerator / (denominator + 1e-10)
  } else {
    ratio <- numerator / denominator
  }
  
  return(ratio)
}


```

## Highlights


## References

## Code Appendix

### To arbitrarily choose r,

```{r to_arbitraily_choose_r}
# Get parameters for source
source_params <- list(
  mu1 = mean(source_data[,1]),
  sigma1 = sd(source_data[,1]),
  mu2 = mean(source_data[,2]),
  sigma2 = sd(source_data[,2])
  )

# Get parameters for target
target_params <- list(
  mu1 = mean(target_data[,1]),
  sigma1 = sd(target_data[,1]),
  mu2 = mean(target_data[,2]),
  sigma2 = sd(target_data[,2])
  )

# How to arbitrarily choose r(How the Kpotufe does)
# Load necessary libraries


# Function to compute true density ratio
# This needs to be defined based on your data generation process
# For example, if both source and target are mixtures of Gaussians:
compute_true_density_ratio <- function(x, source_params, target_params) {
  # x: matrix of points where to compute the ratio
  # source_params and target_params: list containing means and covariances
  # Assuming both source and target are mixtures of two Gaussians

  # Source density
  p_source <- 0.5 * dmvnorm(x, mean = source_params$mu1, sigma = source_params$sigma1) +
              0.5 * dmvnorm(x, mean = source_params$mu2, sigma = source_params$sigma2)

  # Target density
  p_target <- 0.5 * dmvnorm(x, mean = target_params$mu1, sigma = target_params$sigma1) +
              0.5 * dmvnorm(x, mean = target_params$mu2, sigma = target_params$sigma2)

  # Avoid division by zero
  p_source[p_source == 0] <- 1e-10

  # Density ratio
  omega_true <- p_target / p_source
  return(omega_true)
}

# Function to compute histograms and estimate density ratios
estimate_density_ratio_histogram <- function(source_data, target_data, r, d, num_bins) {
  # Create bin edges based on r
  # Assuming data is scaled to [0,1] in each dimension
  bin_edges <- seq(0, 1, by = r)

  # Function to assign data points to bins
  assign_bins <- function(data, bin_edges) {
    apply(data, 2, function(x) cut(x, breaks = bin_edges, include.lowest = TRUE, labels = FALSE))
  }

  # Assign bins
  source_bins <- assign_bins(source_data, bin_edges)
  target_bins <- assign_bins(target_data, bin_edges)

  # Convert multi-dimensional bins to single indices
  source_indices <- apply(source_bins, 1, function(row) paste(row, collapse = "-"))
  target_indices <- apply(target_bins, 1, function(row) paste(row, collapse = "-"))

  # Count occurrences in each bin
  source_counts <- table(factor(source_indices, levels = unique(c(source_indices, target_indices))))
  target_counts <- table(factor(target_indices, levels = unique(c(source_indices, target_indices))))

  # Convert counts to density estimates
  p_source_hat <- as.numeric(source_counts) / length(source_data[,1])
  p_target_hat <- as.numeric(target_counts) / length(target_data[,1])

  # Avoid division by zero
  p_source_hat[p_source_hat == 0] <- 1e-10

  # Estimated density ratio
  omega_hat <- p_target_hat / p_source_hat

  return(list(omega_hat = omega_hat, source_counts = source_counts, target_counts = target_counts))
}

# Main function to select optimal r
select_optimal_r <- function(source_data, target_data, d, n0, n1,
                             beta_samples = 100, num_bins = NULL, seed = 123,
                             source_params = NULL, target_params = NULL) {
  set.seed(seed)

  # Normalize data to [0,1] in each dimension for simplicity
  normalize_data <- function(data) {
    apply(data, 2, function(x) (x - min(x)) / (max(x) - min(x)))
  }

  source_data_norm <- normalize_data(source_data)
  target_data_norm <- normalize_data(target_data)

  # If num_bins is not provided, set it based on r
  # We'll compute r based on beta, so num_bins will be 1/r
  # To ensure integer bins, we'll adjust accordingly

  # Sample beta's uniformly from (0,1]
  beta_values <- runif(beta_samples, min = 1e-3, max = 1 + 1e-15)  # Avoid exactly 0

  # Compute r for each beta
  min_n <- min(n0, n1)
  log_min_n <- log(min_n)

  r_values <- (log_min_n / min_n)^(beta_values / (2 * beta_values + d))

  # Initialize vectors to store bias and variance
  bias_values <- numeric(beta_samples)
  variance_values <- numeric(beta_samples)

  # Loop over each beta and compute bias and variance
  for (i in 1:beta_samples) {
    beta <- beta_values[i]
    r <- r_values[i]

    # Compute number of bins per dimension
    # Assuming r is the bin width, number of bins = ceiling(1 / r)
    bins_per_dim <- ceiling(1 / r)

    # Avoid extremely large number of bins
    if (bins_per_dim > 50) {  # Arbitrary limit
      bins_per_dim <- 50
    }

    # Adjust r based on bins_per_dim
    r_adj <- 1 / bins_per_dim

    # Estimate density ratios using histogram
    hist_est <- estimate_density_ratio_histogram(source_data_norm, target_data_norm, r_adj, d, num_bins = bins_per_dim)
    omega_hat <- hist_est$omega_hat

    # Compute bias and variance
    # To compute bias and variance, we need the true omega at each bin
    # We'll approximate the bin centers
    # Generate all possible bin centers
    grid_list <- replicate(d, seq(r_adj/2, 1 - r_adj/2, by = r_adj), simplify = FALSE)
    grid <- expand.grid(grid_list)

    # Compute true omega at each bin center
    omega_true <- compute_true_density_ratio(as.matrix(grid), source_params, target_params)

    # Compute estimated omega at each bin
    # Map histogram omega_hat to grid
    # Assuming the bins are ordered consistently
    # Compute bias: |omega_hat - omega_true|
    bias <- mean(abs(omega_hat - omega_true))

    # Compute variance: variance of omega_hat
    variance <- var(omega_hat)

    bias_values[i] <- bias
    variance_values[i] <- variance
  }

  # Find the r where bias is approximately equal to variance
  # Compute the absolute difference between bias and variance
  diff_bv <- abs(bias_values - variance_values)

  # Find the index with the minimum difference
  optimal_index <- which.min(diff_bv)

  optimal_r <- r_values[optimal_index]
  optimal_beta <- beta_values[optimal_index]
  optimal_bias <- bias_values[optimal_index]
  optimal_variance <- variance_values[optimal_index]

  # Optionally, plot bias and variance vs r
  plot_df <- data.frame(
    r = r_values,
    bias = bias_values,
    variance = variance_values,
    diff = diff_bv
  )

  p1 <- ggplot(plot_df, aes(x = r)) +
    geom_line(aes(y = bias, color = "Bias")) +
    geom_line(aes(y = variance, color = "Variance")) +
    geom_vline(xintercept = optimal_r, linetype = "dashed", color = "black") +
    labs(title = "Bias and Variance vs r",
         x = "r",
         y = "Value",
         color = "") +
    theme_minimal()

  p2 <- ggplot(plot_df, aes(x = r, y = diff)) +
    geom_point(alpha = 0.6) +
    geom_vline(xintercept = optimal_r, linetype = "dashed", color = "black") +
    labs(title = "Difference between Bias and Variance vs r",
         x = "r",
         y = "|Bias - Variance|") +
    theme_minimal()

  grid.arrange(p1, p2, ncol = 2)

  # Return the optimal r and related information
  return(list(
    optimal_r = optimal_r,
    optimal_beta = optimal_beta,
    bias = optimal_bias,
    variance = optimal_variance,
    all_r = r_values,
    all_bias = bias_values,
    all_variance = variance_values
  ))
}

# Example Usage

# Define source and target parameters for true density ratio computation
# Example: both are mixtures of two Gaussians
source_params <- list(
  mu1 = c(0, 0),
  sigma1 = diag(2),
  mu2 = c(3, 3),
  sigma2 = diag(2)
)

target_params <- list(
  mu1 = c(1, 1),
  sigma1 = diag(2) * 0.5,
  mu2 = c(-2, -2),
  sigma2 = diag(2) * 2
)

# Simulate data using the previously defined simulate_data function
simulate_data <- function(n0 = 1000, n1 = 1000, d = 2, seed = 123) {
  set.seed(seed)

  # Source Data
  source_data <- rbind(
    mvrnorm(n = n0 / 2, mu = source_params$mu1, Sigma = source_params$sigma1),
    mvrnorm(n = n0 / 2, mu = source_params$mu2, Sigma = source_params$sigma2)
  )

  # Target Data
  target_data <- rbind(
    mvrnorm(n = n1 / 2, mu = target_params$mu1, Sigma = target_params$sigma1),
    mvrnorm(n = n1 / 2, mu = target_params$mu2, Sigma = target_params$sigma2)
  )

  list(source = source_data, target = target_data)
}

# Generate simulated data
sim_data <- simulate_data(n0 = 1000, n1 = 1000, d = 2, seed = 123)
source_data <- sim_data$source
target_data <- sim_data$target

# Select optimal r
optimal_selection <- select_optimal_r(source_data, target_data, d = 2,
                                     n0 = 1000, n1 = 1000,
                                     beta_samples = 100, num_bins = NULL, seed = 123,
                                     source_params = source_params,
                                     target_params = target_params)


```

Here we want to know how the approximate interval that $r$ would fall into. So we simulate the data-driven method used in [Kpotufe(2017)](https://proceedings.mlr.press/v54/kpotufe17a/kpotufe17a.pdf).

With the simulation, we have $r$ = `r optimal_selection$optimal_r`, $\beta$ = `r optimal_selection$optimal_beta`. We choose the $r$ when its variance(= `r optimal_selection$variance`) and bias(= `r optimal_selection$bias`) term are approximately equal.