---
title: "Kullback-Leibler method"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Theoretical Background
Inspired by the idea that MLE is equivalent to minimizing KL divergence:

$KL(\mathbb{P}^{(0)}||\tilde{\mathbb{P}}^{(0)}) = \mathbb{E}_{X \sim \mathbb{P}^{(0)}}[log(\frac{d\mathbb{P}^{(0)}}{d\mathbb{P}^{(1)}}(X))] - \mathbb{E}_{X \sim \mathbb{P}^{(0)}}[log \hat{w}(X)]$

We can estimate the density ratio $\hat{w}$ by minimizing KL divergence $D_{KL}$ (maximizing $\mathbb{E}_{X \sim \mathbb{P}^{(0)}}[log \hat{w}(X)]$).

Estimator:

$\hat{w} = \arg\max_{h: X \to \mathbb{R}_{+}}\{\mathbb{E}_{X \sim \hat{\mathbb{P}}^{(0)}}[log h(X)] - \mathbb{E}_{X \sim \hat{\mathbb{P}}^{(1)}}[h(X)]\}$

In practice, we use a specific function class $\mathcal{G}$ (e.g. RKHS with a universal kernel)
to estimate $h: X \to \mathbb{R}_{+}$

# Implementation

```{r}
# To install 'densityratio' package, we need libraries:
library(osqp)
library(pbapply)
library(RcppProgress)
#install.packages('densityratio', repos = 'https://thomvolker.r-universe.dev')
library(densityratio)

head(numerator_data)


fit_kliep <- kliep(
  df_numerator = numerator_data$x5, 
  df_denominator = denominator_data$x5
)

summary(fit_kliep)
# Predict estimated density ratio
pred_kliep  <- predict(fit_kliep, newdata = denominator_data$x5)
summary(pred_kliep)
# Function of true density ratio
dbinorm <- function(x, p, dif) {
  p * dnorm(x, 0, 1) + (1 - p) * dnorm(x, dif, 1)
}
dbinorm_scaled <- function(x, scaling, ...) {
  dbinorm(x, ...) * scaling
}
dnorm_scaled <- function(x, scaling, ...) {
  dnorm(x, ...) * scaling
}

dratio_kliep <- function(x, p, dif, mu, sd) {
  dbinorm(x, p, dif) / dnorm(x, mu, sd)
}

```

```{r}
ggplot() + 
  geom_density(aes(x = numerator_data$x5, colour = "target")) +
  geom_density(aes(x = denominator_data$x5, colour = "source"))

ggplot() +
  geom_density(aes(x = numerator_data$x5, color = "target")) +
  geom_density(aes(x = denominator_data$x5, weight = as.numeric(pred_kliep), color = "weighted source"))


ggplot() +
  geom_point(aes(x = denominator_data$x5, y = pred_kliep, col = "estimates")) +
  stat_function(mapping = aes(col = "True density ratio"), 
                fun = dratio_kliep, 
                args = list(p = 0.4, dif = 3, mu = 3, sd = 2),
                linewidth = 1)
  
```

```{r cache=TRUE}
dat <- read.csv("./data/data_forSDE.csv")
library(tidyr)
library(dplyr)


# univariate data
source_data1 <- dat %>%
  filter(mechvent == 1, los<150) %>%
  dplyr::select(los) %>%
  drop_na()

target_data1 <- dat %>%
  filter(mechvent == 0,los<150) %>%
  dplyr::select(los) %>%
  drop_na()

# multivariate data
source_data3 <- dat %>%
  filter(mechvent == 1,los<150) %>%
  dplyr::select(los, heartrate, age) %>%
  drop_na()

target_data3 <- dat %>%
  filter(mechvent == 0,los<150) %>%
  dplyr::select(los, heartrate, age) %>%
  drop_na()

fit_kliep1 <- kliep(
  df_numerator = target_data1, 
  df_denominator = source_data1,
  nsigma = 10, ncenters = 150, nfold = 5,
  epsilon = 10^{3:-5}, maxit = 1000
)

fit_kliep3 <- kliep(
  df_numerator = target_data3, 
  df_denominator = source_data3,
  nsigma = 20, ncenters = 100, nfold = 5,
  epsilon = 10^{3:-5}, maxit = 1000
  )

pred1_kliep <- predict(fit_kliep1, newdata = source_data1)
pred3_kliep <- predict(fit_kliep3, newdata = source_data3)
```

```{r}
ggplot() +
  geom_density(aes(x = los, color = "Source: mechvent=1"), data = source_data1, size = 1) +
  geom_density(aes(x = los, color = "Target: mechvent=0"), data = target_data1, size = 1) +
  labs(title = "LOS Distribution Shift by MechVent Status", x = "Length of Stay", y = "Density") +
  scale_color_manual(values = c("Source: mechvent=1" = "blue", "Target: mechvent=0" = "red")) +
  theme_minimal()


ggplot() +
  geom_density(aes(x = los, weight = pred1, color = "Weighted Source (mechvent=1)"), data = source_data1, size = 1) +
  geom_density(aes(x = los, color = "Target (mechvent=0)"), data = target_data1, size = 1) +
  scale_color_manual(values = c("Weighted Source (mechvent=1)" = "blue", "Target (mechvent=0)" = "red")) +
  labs(title = "LOS: Weighted Source vs. Target (MechVent)", x = "Length of Stay", y = "Density") +
  theme_minimal()
```

```{r}
# Marginal densities of los
ggplot() +
  geom_density(aes(x = los, color = "Target(mechvent=0)", fill = "Target(mechvent=0)"), 
               data = target_data3, alpha = 0.3, size = 1) +
  geom_density(aes(x = los, weight = pred3_kliep, color = "Weighted Source(mechvent=1)", fill = "Weighted Source(mechvent=1)"), 
               data = source_data3, alpha = 0.3, size = 1) +
  scale_color_manual(name = "Dataset", values = c("Weighted Source(mechvent=1)" = "blue", "Target(mechvent=0)" = "red")) +
  scale_fill_manual(name = "Dataset", values = c("Weighted Source(mechvent=1)" = "blue", "Target(mechvent=0)" = "red")) +
  labs(title = "Density of Univariate Data(LOS): Weighted Source vs. Target",
       x = "Length of Stay", y = "Density") +
  theme_minimal()

ggplot() +
  geom_density(aes(x = age, color = "Target(mechvent=0)", fill = "Target(mechvent=0)"), 
               data = target_data3, alpha = 0.3, size = 1) +
  geom_density(aes(x = age, weight = pred3_kliep, color = "Weighted Source(mechvent=1)", fill = "Weighted Source(mechvent=1)"), 
               data = source_data3, alpha = 0.3, size = 1) +
  scale_color_manual(name = "Dataset", values = c("Weighted Source(mechvent=1)" = "blue", "Target(mechvent=0)" = "red")) +
  scale_fill_manual(name = "Dataset", values = c("Weighted Source(mechvent=1)" = "blue", "Target(mechvent=0)" = "red")) +
  labs(title = "Density of Univariate Data(Age): Weighted Source vs. Target",
       x = "Length of Stay", y = "Density") +
  theme_minimal()


ggplot() +
  geom_density(aes(x = heartrate, color = "Target(mechvent=0)", fill = "Target(mechvent=0)"), 
               data = target_data3, alpha = 0.3, size = 1) +
  geom_density(aes(x = heartrate, weight = pred3_kliep, color = "Weighted Source(mechvent=1)", fill = "Weighted Source(mechvent=1)"), 
               data = source_data3, alpha = 0.3, size = 1) +
  scale_color_manual(name = "Dataset", values = c("Weighted Source(mechvent=1)" = "blue", "Target(mechvent=0)" = "red")) +
  scale_fill_manual(name = "Dataset", values = c("Weighted Source(mechvent=1)" = "blue", "Target(mechvent=0)" = "red")) +
  labs(title = "Density of Univariate Data(Heart Rate): Weighted Source vs. Target",
       x = "Length of Stay", y = "Density") +
  theme_minimal()



# Reduce dimension w/ PCA
# For the source data, include the weight; for target, set weight = 1.
source_data3 <- source_data3 %>% mutate(dataset = "source", weight = pred3_kliep)
target_data3 <- target_data3 %>% mutate(dataset = "target", weight = 1)

combined_data3 <- bind_rows(source_data3, target_data3)

# Run PCA on the three variables: age, los, and heartrate.
pca_res <- prcomp(combined_data3[, c("age", "los", "heartrate")], scale. = TRUE)
combined_data3 <- combined_data3 %>%
  mutate(PC1 = pca_res$x[,1],
         PC2 = pca_res$x[,2])

# Plot density contours for the target distribution.
ggplot() +
  geom_density_2d(data = combined_data3 %>% filter(dataset == "target"), 
                    aes(x = PC1, y = PC2, color = "Target(mechvent=0)"), linetype = "solid", size = 1) +
  # Overlay a weighted density estimate for the source.
  stat_density_2d(data = combined_data3 %>% filter(dataset == "source"), 
                    aes(x = PC1, y = PC2, weight = weight, color = "Weighted Source(mechvent=1)"), 
                    geom = "density_2d", linetype = "dashed", size = 1) +
  labs(title = "PCA of Multivariate Data: Weighted Source vs. Target",
       x = "PC1", y = "PC2") +
  theme_minimal()

```

# Highlights

We use R packages 'densityratio' which contains sample data and kernel-based density ratio estimation methods.

# References

- https://github.com/thomvolker/densityratio.git
