---
title: "Kullback-Leibler method"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Theoretical Background
Inspired by the idea that MLE is equivalent to minimizing KL divergence:

$KL(\mathbb{P}^{(0)}||\tilde{\mathbb{P}}^{(0)}) = \mathbb{E}_{X \sim \mathbb{P}^{(0)}}[log(\frac{d\mathbb{P}^{(0)}}{d\mathbb{P}^{(1)}}(X))] - \mathbb{E}_{X \sim \mathbb{P}^{(0)}}[log \hat{w}(X)]$

We can estimate the density ratio $\hat{w}$ by minimizing KL divergence $D_{KL}$ (maximizing $\mathbb{E}_{X \sim \mathbb{P}^{(0)}}[log \hat{w}(X)]$).

Estimator:

$\hat{w} = \arg\max_{h: X \to \mathbb{R}_{+}}\{\mathbb{E}_{X \sim \hat{\mathbb{P}}^{(0)}}[log h(X)] - \mathbb{E}_{X \sim \hat{\mathbb{P}}^{(1)}}[h(X)]\}$

In practice, we use a specific function class $\mathcal{G}$ (e.g. RKHS with a universal kernel)
to estimate $h: X \to \mathbb{R}_{+}$

# Implementation

```{r}
# To install 'densityratio' package, we need libraries:
library(osqp)
library(pbapply)
library(RcppProgress)
#install.packages('densityratio', repos = 'https://thomvolker.r-universe.dev')
library(densityratio)

head(numerator_data)


fit_kliep <- kliep(
  df_numerator = numerator_data$x5, 
  df_denominator = denominator_data$x5
)

summary(fit_kliep)
# Predict estimated density ratio
pred_kliep  <- predict(fit_kliep, newdata = denominator_data$x5)
summary(pred_kliep)
# Function of true density ratio
dbinorm <- function(x, p, dif) {
  p * dnorm(x, 0, 1) + (1 - p) * dnorm(x, dif, 1)
}
dbinorm_scaled <- function(x, scaling, ...) {
  dbinorm(x, ...) * scaling
}
dnorm_scaled <- function(x, scaling, ...) {
  dnorm(x, ...) * scaling
}

dratio_kliep <- function(x, p, dif, mu, sd) {
  dbinorm(x, p, dif) / dnorm(x, mu, sd)
}

```

```{r}
ggplot() + 
  geom_density(aes(x = numerator_data$x5, colour = "target")) +
  geom_density(aes(x = denominator_data$x5, colour = "source"))

ggplot() +
  geom_density(aes(x = numerator_data$x5, color = "target")) +
  geom_density(aes(x = denominator_data$x5, weight = as.numeric(pred_kliep), color = "weighted source"))


ggplot() +
  geom_point(aes(x = denominator_data$x5, y = pred_kliep, col = "estimates")) +
  stat_function(mapping = aes(col = "True density ratio"), 
                fun = dratio_kliep, 
                args = list(p = 0.4, dif = 3, mu = 3, sd = 2),
                linewidth = 1)
  
```



# Highlights

We use R packages 'densityratio' which contains sample data and kernel-based density ratio estimation methods.

# References

- https://github.com/thomvolker/densityratio.git
