---
title: "Least Square Method"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Theoretical Background

Intuition: 

- ERM $E_{(X,Y)\sim \mathbb{P}^{(0)}}[l(\hat{h}(X), Y)] \leq min E_{(X,Y)\sim \mathbb{P}^{(1)}}[w(X)l(h(X), Y)] + O_{\mathbb{P}}(1) + 2CE_{X\sim \hat{\mathbb{P}^{(1)}}}|\hat{w}(X) - w(X)|$

- Goal: find $\hat{w}(X)$ that minimizes the cost of weight estimation term $2CE_{X\sim \hat{\mathbb{P}^{(1)}}}|\hat{w}(X) - w(X)|$

Estimator:

- Formulate our goal into: 

  $min_{\alpha \in R^b} \frac{1}{2} \alpha^{T} \hat{H}\alpha - \hat{h}^{T} \alpha + \lambda 1_{b}^{T} \alpha$

  $s.t. \alpha \geq 0_{b}$ in constrained case

By [Kanamori et al. (2009)](), unconstrained method performed better than the constrained version in practice; there is a kernelized version w/ ridge penalty term as well.




# Implementation

```{r}
# To install 'densityratio' package, we need libraries:
library(osqp)
library(pbapply)
library(RcppProgress)
#install.packages('densityratio', repos = 'https://thomvolker.r-universe.dev')
library(densityratio)

head(numerator_data)


fit_ulsif <- ulsif(
  df_numerator = numerator_data$x5, 
  df_denominator = denominator_data$x5
)

summary(fit_ulsif)
# Predict estimated density ratio
pred_ulsif  <- predict(fit_ulsif, newdata = denominator_data$x5)
summary(pred_ulsif)
# Function of true density ratio
dbinorm <- function(x, p, dif) {
  p * dnorm(x, 0, 1) + (1 - p) * dnorm(x, dif, 1)
}
dbinorm_scaled <- function(x, scaling, ...) {
  dbinorm(x, ...) * scaling
}
dnorm_scaled <- function(x, scaling, ...) {
  dnorm(x, ...) * scaling
}

dratio_ulsif <- function(x, p, dif, mu, sd) {
  dbinorm(x, p, dif) / dnorm(x, mu, sd)
}

```

```{r}
library(ggplot2)
ggplot() + 
  geom_density(aes(x = numerator_data$x5, colour = "target")) +
  geom_density(aes(x = denominator_data$x5, colour = "source"))

ggplot() +
  geom_density(aes(x = numerator_data$x5, color = "target")) +
  geom_density(aes(x = denominator_data$x5, weight = as.numeric(pred_ulsif), color = "weighted source"))


ggplot() +
  geom_point(aes(x = denominator_data$x5, y = pred_ulsif, col = "estimates")) +
  stat_function(mapping = aes(col = "True density ratio"), 
                fun = dratio_ulsif, 
                args = list(p = 0.4, dif = 3, mu = 3, sd = 2),
                linewidth = 1)
  
```

# Highlights

# Reference

- https://github.com/thomvolker/densityratio.git