---
title: "Semi-parametric Method"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Theoretical Background

- Profile empirical log-likelihood ratio function: $l(\theta, \lambda) = - \sum_{k=0}^{1} \sum_{i=1}^{n_k} log [1 + \lambda(g(x^{(k)}_i ; \theta) + 1)] + \sum_{i=1}^{n_1} log g(x^{(1)}_i ; \theta)$


- Estimation:

  **step1**: Let $\frac{dl}{d\theta} = \frac{dl}{d\lambda} = 0 \Rightarrow (\hat{\theta}. \hat{\lambda})$

  **step2**: Estimate density ratio: $\hat{w} = \hat{g}(x; \hat{\theta})$

When implementing step1 in practice, the estimated parameters are very likely to be `Inf` due to the log term and exponential tilting link function $g(x; \theta)$ Profile log-likelihood function.

# Implementation

```{r}
# To load 'densityratio', we need 3 libraries:
library(osqp)
library(pbapply)
library(RcppProgress)
#install.packages('densityratio', repos = 'https://thomvolker.r-universe.dev')
library(densityratio)
##############################################################
# Example: Exponential Tilting to Estimate a Density Ratio
##############################################################
# X_target <- numerator_data$x5
# X_source <- denominator_data$x5
# (1) Suppose we have 1D data
set.seed(123)

n_source <- 500
n_target <- 500

X_source <- rbeta(n_source, shape1 = 2, shape2 = 2)
X_target <- runif(n_target, min = 0, max = 1)


combined_df <- data.frame(
  target = X_target, 
  source = X_source
)

# (2) Define the link function g(x; beta) = exp(beta0 + beta1*x).
g_fun <- function(x, param) {
  beta1 <- param[1]
  beta2 <- param[2]
  alpha <- param[3]
  # you could also add more terms or expansions of x
  return(exp(alpha + beta1 * log(x) + beta2 * log(1 - x)) )
}

# (3) Define the objective function 
# We'll implement the NEGATIVE of this, so we can use 'optim' (which by default MINIMIZES the objective).
loglik_qin <- function(param, X_source, X_target) {
  lambda <- param[4]
  
  # Possibly print for debugging:
  # cat("beta0=",beta0,"beta1=",beta1,"lambda=",lambda,"\n")
  
  g_target <- g_fun(X_target, param)
  g_source <- g_fun(X_source, param)

  # Check domain
  arg_target <- 1 + lambda * (g_target + 1)
  arg_source <- 1 + lambda * (g_source + 1)
  
  if (any(!is.finite(arg_target)) || any(!is.finite(arg_source))) {
  # We have Inf or NaN in arg_target or arg_source
  return(Inf)
}

  if (any(arg_target <= 0) || any(arg_source <= 0)) {
    # cat("Invalid log argument\n")
    return(Inf)
  }

  # Then do your sums:
  eps <- 1e-12
  term_target       <- sum(log(arg_target + eps))
  term_source_part1 <- sum(log(arg_source + eps))
  term_source_part2 <- sum(log(g_source + eps))
  
  # Negative log-likelihood or similar
  val <-  term_target + term_source_part1 - term_source_part2

  # Check if final is finite
  if(!is.finite(val)) {
    # cat("Non-finite final val!\n")
    return(Inf)
  }

  return(val)
}


init_param <- c(beta1 = 1, beta2 = 1, alpha = 1.792, lambda = 0.03)

val <- loglik_qin(init_param, X_source, X_target)

cat("Function value at initial guess:", val, "\n")
cat("Is it finite?", is.finite(val), "\n")
```




```{r}

# (4) We'll define a function to do the optimization
qin_fit <- function(X_source, X_target) {
  init_param <- c(beta1 = 1, beta2 = 1, alpha = 1.792, lambda = 0.03)
  
  fit <- optim(
    par       = init_param,
    fn        = loglik_qin,  # Minimizing NEG log-likelihood
    X_source  = X_source,
    X_target  = X_target,
    method    = "BFGS",
    control   = list(maxit=1),
    hessian   = FALSE
  )
  
  return(fit)
}

# (5) Run the fit
fit_result <- qin_fit(X_source, X_target)
fit_result

# (6) Extract the parameter estimates
beta1_hat <- fit_result$par[1]
beta2_hat <- fit_result$par[2]
alpha_hat <- fit_result$par[3]


# (7) Define the estimated ratio function w_hat(x)
w_hat <- function(x) {
  exp(alpha_hat + beta1_hat * log(x) + beta2_hat * log(1-x))  # g_fun(x, c(beta0_hat, beta1_hat))
}
w <- exp(1.792 + 1 * log(X_source) + 1 * log(1-X_source))
# Now you can use w_hat(x) to reweight X_source or for any purpose.
weights <- w_hat(X_source)

```

```{r}
library(ggplot2)
ggplot() + 
  geom_density(aes(x = X_target, colour = "target")) +
  geom_density(aes(x = X_source, colour = "source"))

ggplot() +
  geom_density(aes(x = X_target, color = "target")) +
  geom_density(aes(x = X_source, weight = weights, color = "weighted source"))

ggplot() +
  geom_density(aes(x = X_target, color = "target")) +
  geom_density(aes(x = X_source, weight = w, color = "weighted source"))
```

